---
title: "Paper 4"
author: "Niklas Karlsen"
date: "2024-08-22"
output: 
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    
knit: (function(inputFile, encoding) { rmarkdown::render(inputFile, encoding = encoding, output_file = file.path(dirname(inputFile), '../output/paper4.html')) })
---

This is a computational essay?

Changelog

* 22.8.2024: added comparison of two hybrid models

TODO

# Preamble / technical details

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE) #,results="asis")

Sys.setenv(LANG = "no_NB.UTF-8")            # for æ,ø og å. Kilde:
Sys.setlocale("LC_ALL","Norwegian")
# https://stackoverflow.com/questions/16347731/how-to-change-the-locale-of-r

# variable
plot_to_file = 0
```

```{r last_bibliotek, echo=FALSE, warning=FALSE, message=FALSE}
## Last bibliotek: rekkefølge - de siste overstyrer de første

## bibliotek for tabell-behandling
# last plyr før dplyr
#library(plyr)                               # for plot of mean and sd
library(dplyr)
library(tidyr)
library(pander)                             # prettyprint for tables
# kilde: https://stackoverflow.com/questions/31626419/in-knitr-no-output-from-pander-in-for-loop
#panderOptions('knitr.auto.asis', FALSE)     # to pander in loop
#library(tidyverse)

## bibliotek for plotting
#library(plotrix)                            # SE(m) (plot mean+CI)
library(ggplot2)
#library(GGally)                             # for plot av ggpairs
# kilde: https://ggplot2-book.org/arranging-plots.html#arranging-plots
#library(patchwork)                          # for inset_element

## bibliotek for statistikk
#library(bruceR,quietly=TRUE)                           # for manova
#library(DescTools)                          # for lambda, categorical data
library(psych)                              # for ICC, SD
library(effsize)                            # for cohen's d
#library(corrplot)
#library(rcompanion)                          # for Cramers V
#library(Rfit)                               # ranked regression # https://journal.r-project.org/archive/2013/RJ-2013-027/RJ-2013-027.pdf

### regression libraries
library(lme4)                                 # for MLM
#library(lmerTest)                             # not recommended? Avoid the lmerTest package (https://utstat.toronto.edu/~brunner/workshops/mixed/NormalWithR.pdf)
library(car)                                   # # for p-values i lme4 (anova, test="F"). durbinWatson
library(nlme)                                 # for lme
#library(rmcorr)
library(plm)                                  # paneldata analysis

# t-tests
library(WRS2)                                 # Yuen-Welch test ref.: Mair & Wilcox, 2020

# SE calculation
library(pcse)                                 # panel-corrected SE
library(sandwich) # cluster corrected error
library(lmtest) # logLikelihood tests, AIC, BIC ,...
#library(MuMIn) # AICc
library(aod) # wald.test

library(texreg)                               # screenreg (print compare results regresion)
library(viridis)

#library(ppcor) # partiell correlation
#library(mice) # missing
```

## Definer noen hjelpefunksjoner.

```{r last_hjelpefunksjoner}
source("../../scripts/help-functions.R")
```

## Definer constructs for lettere access i kolonnene

```{r load_constructs}
source("../../scripts/constructs.R")
```

# Introduction

The purpose of this script is to do the calculations presented in paper 4 for prototype 1 and 2.

* Difference (gain) scores / Hedges' g
* Fixed-effects analysis
* Within-between (hybrid) model

# Methods

## Samples

Properties of samples

Load data

```{r load_data}
# Rdata-file contains reduced / prepared datasets
load(file="../../data/processed_data/tpack-pilot_2022.Rdata")
load(file="../../data/processed_data/tpack-emne2_2023.Rdata")

# fix bug: data_emne2_ren is list not dataframe
data_emne2_ren <- data.frame(data_emne2_ren)
```

Synchronize data structures / naming conventions

```{r create_dataframe_v23}
# [c(1:4,6)] / unlist(TPACK_PST_emne2_v23)
data_emne2_v23 <- data_emne2_ren[,colnames(data_emne2_ren) %in% unlist(TPACK_PST_emne2_v23)]  #- 3 

#transfer covariates
data_emne2_v23$Gender   <- c(data_pre1_emne2$Gender, 
                             rep(NA, nrow(data_pre2_emne2)),
                             data_post_emne2$Gender) # data_emne2_ren$Gender
data_emne2_v23$Trinn    <- c(data_pre1_emne2$Trinn, 
                             rep(NA, nrow(data_pre2_emne2)),
                             data_post_emne2$Trinn) # data_emne2_ren$Trinn
data_emne2_v23$praksis  <- c(data_pre1_emne2$'Har du brukt programmering i praksis sammen med skoleelever?', 
                             rep(NA, nrow(data_pre2_emne2)),
                             data_post_emne2$'Har du brukt programmering i praksis sammen med skoleelever?') 

data_emne2_v23$full_treatment <- rep(0,nrow(data_emne2_v23))

data_emne2_v23$datasett <- data_emne2_ren$datasett
data_emne2_v23$Subject  <- c(data_pre1_emne2$Subject, 
                             data_pre2_emne2$Subject,
                             data_post_emne2$Subject)
```

sjekk at full_treatment blir registert

```{r}
# denne måtte skilles ut for å bli registrert riktig!?
data_emne2_v23$full_treatment[data_emne2_v23$Subject %in% data_post_emne2$Subject] <- 1

table(data_emne2_v23$full_treatment)
```

```{r create_dataframe_v22}

# make working copy
data_emne2_v22    <-  data_TPACK_pilot  [data_TPACK_pilot$datasett == "pre" | 
                                           data_TPACK_pilot$datasett == "post",
                                         !colnames(data_TPACK_pilot) %in%
                                           c("semester","studiested",
                                             "XK1","XK2","XK3","XK4","XK5","XK6","CK123","CK6")] 

```

Sort datasets

```{r sort_students}
# sort students
# sorter på person
data_emne2_v22$datasett <- factor(data_emne2_v22$datasett, level=c("pre","post"))
data_emne2_v22 <-  data_emne2_v22[
                          with(data_emne2_v22, 
                          order(Subject,datasett, # rekkefølge
                          method="radix", # for faktorer, brukes med c nedenfor
                          decreasing = c(0,0))),] # bare datasett (1) 

# set correct order of dataset
data_emne2_v23$datasett <- factor(data_emne2_v23$datasett, level=c("pre1","pre2","post"))
data_emne2_v23 <-  data_emne2_v23[
                          with(data_emne2_v23, 
                          order(Subject,datasett, # rekkefølge
                          method="radix", # for faktorer, brukes med c nedenfor
                          decreasing = c(0,0))),] # bare datasett (1) 
```

add group information

```{r add_groups_v22}
# add column with group numbers
data_emne2_v22$group <- c(1:nrow(data_emne2_v22))
data_emne2_v22[data_emne2_v22$Subject == 'Gabriel' |
                 data_emne2_v22$Subject == 'Hulda' |
                 data_emne2_v22$Subject == 'Isak' |
                 data_emne2_v22$Subject == 'Maria' |
                 data_emne2_v22$Subject == 'Nicoline',"group"] <-
                 "armageddon" #1

data_emne2_v22[data_emne2_v22$Subject == 'Conny' |
                 data_emne2_v22$Subject == 'Fride' |
                 data_emne2_v22$Subject == 'Jonas' |
                 data_emne2_v22$Subject == 'Julia' |
                 data_emne2_v22$Subject == 'Lea',"group"] <-
                 "fasetilstander" # 2

data_emne2_v22[data_emne2_v22$Subject == 'Berta' |
                 data_emne2_v22$Subject == 'Dina' |
                 data_emne2_v22$Subject == 'Gloria' |
                 data_emne2_v22$Subject == 'Klaudia',"group"] <- 
                 "atomeksitasjon" # 3

data_emne2_v22[data_emne2_v22$Subject == 'Benjamin' |
                 data_emne2_v22$Subject == 'Ester' |
                 data_emne2_v22$Subject == 'Harald' |
                 data_emne2_v22$Subject == 'Inga' |
                 data_emne2_v22$Subject == 'Olava',"group"] <- 
                 "smittespredning" # 4

data_emne2_v22[data_emne2_v22$Subject == 'David' |
                 data_emne2_v22$Subject == 'Elias' |
                 data_emne2_v22$Subject == 'Filip',"group"] <-
                 "trykkbeholder" # 5

data_emne2_v22 <- data_emne2_v22 %>% relocate(group, .after = Subject)           # bruker dplyr
```

```{r add_groups_v23}
# add column with group numbers
data_emne2_v23$group <- rep(NA,nrow(data_emne2_v23))
data_emne2_v23[data_emne2_v23$Subject == 'Berta' |
                 data_emne2_v23$Subject == 'Fride' |
                 data_emne2_v23$Subject == 'Gloria' |
                 data_emne2_v23$Subject == 'Nora',"group"] <-
                 "kokeplate" #1/A

data_emne2_v23[data_emne2_v23$Subject == 'Conny' |
                 data_emne2_v23$Subject == 'Idun' |
                 data_emne2_v23$Subject == 'Magrit' |
                 data_emne2_v23$Subject == 'Vaarin',"group"] <-
                 "arv øyefarge" # 2/B

data_emne2_v23[data_emne2_v23$Subject == 'Julia' |
                 data_emne2_v23$Subject == 'Dina' |
                 data_emne2_v23$Subject == 'Ruben',"group"] <- 
                 "partikkelmodellen gass" # 3/C

data_emne2_v23[data_emne2_v23$Subject == 'Emmanuel' |
                 data_emne2_v23$Subject == 'Klaudia' |
                 data_emne2_v23$Subject == 'Noah' | 
                 data_emne2_v23$Subject == "Christian","group"] <- 
                 "partikkelmodellen vannglass" # 4/E

data_emne2_v23[data_emne2_v23$Subject == 'Gabriel' |
                 data_emne2_v23$Subject == 'Harald' |
                 data_emne2_v23$Subject == 'Lukas' |
                 data_emne2_v23$Subject == 'Mikael',"group"] <-
                 "friksjon" # 5/F

data_emne2_v23[data_emne2_v23$Subject == 'Olava' |
               data_emne2_v23$Subject == "Yosita" | 
               data_emne2_v23$Subject == "Nicoline" ,"group"] <-
                 "faseoverganger grønn rute" # 6/D

data_emne2_v23 <- data_emne2_v23 %>% relocate(group, .after = Subject)           # bruker dplyr
```

## Instruments (data collection)

Data were collected using the measure developed by Karlsen et al. (2024) / paper 1. 

The measure used in paper 4 was an early version of the instrument developed in paper 1. This led to the TK-items being somewhat different from the ones in paper 1. Here I use TK2, TK5 and TK6, while in paper 1 I use TK1, TK2 and TK3. 

## Analyses

For easier comparison between the different analysis methods, the same sample is used for the analyses (i.e., n=19 for prototype1 and n=13 for prototype 2).

Select students

```{r select_students_v22}
###
# v22
###

# bare se på studenter som er med både pre og post
stud_alle_v22  <- intersect(data_emne2_v22$Subject[data_emne2_v22$datasett == "pre"],
                         data_emne2_v22$Subject[data_emne2_v22$datasett == "post"])

# hvilke studenter var ikke med på pre, men på post? 
stud_bare_pre_eller_post_v22  <- data_emne2_v22$Subject[!data_emne2_v22$Subject %in%
                                                          stud_alle_v22]

# hvilke var med bare på post?
stud_bare_post_v22 <- data_emne2_v22$Subject[
  data_emne2_v22$datasett == "post" &
    data_emne2_v22$Subject %in% stud_bare_pre_eller_post_v22]

data_emne2_v22$testing <- rep(0,nrow(data_emne2_v22))
data_emne2_v22$testing[data_emne2_v22$Subject == stud_bare_post_v22] <- 1

alle_studenter_v22 <- data_emne2_v22$Subject %in% stud_alle_v22
```

```{r select_students_v23} 
###
# v23
###

# bare se på studenter som er med både pre og post
stud_alle_v23  <- intersect(data_emne2_v23$Subject[data_emne2_v23$datasett == "pre1"],
                         data_emne2_v23$Subject[data_emne2_v23$datasett == "pre2"])

# hvilke var med bare på pre2?
stud_bare_pre2_v23 <- data_emne2_v23$Subject[
  data_emne2_v23$datasett == "pre2" &
    !data_emne2_v23$Subject %in% stud_alle_v23]

stud_alle_v23  <- intersect(stud_alle_v23,
                         data_emne2_v23$Subject[data_emne2_v23$datasett == "post"])

# hvilke var med bare på post?
stud_bare_post_v23 <- data_emne2_v23$Subject[
  data_emne2_v23$datasett == "post" &
    !data_emne2_v23$Subject %in% stud_alle_v23]

data_emne2_v23$testing <- rep(0,nrow(data_emne2_v23))
data_emne2_v23$testing[data_emne2_v23$Subject %in% stud_bare_pre2_v23] <- 1

#data_emne2_v22 <- data_emne2_v22[data_emne2_v22$Subject %in% stud_alle_v22,]
#data_emne2_v23 <- data_emne2_v23[data_emne2_v23$Subject %in% stud_alle_v23,]

alle_studenter_v23 <- data_emne2_v23$Subject %in% stud_alle_v23
```

### Difference scores

To decide which statistical tests we should use to compare pre-post scores, the Levene’s (Mean) test for equal variances and the Kolmogorov-Smirnov test for normality were used as preliminary tests (Pearce and Derrick, 2019). Due to the small sample sizes, the data are probably non-normally distributed. If the variance of the pre-post differences are unequal, this suggests the use of the non-parametric Yuen d-test with trimmed means (Mair & Wilcox, 2020), which show good power for low samples (Wanishsakpong et al., 2023) and can be appropriate when one is not interested in generalizing to a population (Ng  & Cribbie, 2019). If, on the other hand, the pre-post data have equal variance, this suggests the use of the Mann-Whitney (i.e., Wilcox) test with arithmetic mean.

Hedges' g was also calculated, since it is often used in meta-analyses.

#### Prototype 1

```{r descriptive_stat_v22}
scores_v22 <- scoreItems(TPACK_PST_emne2_v22, data_emne2_v22)
data_emne2_v22$TK <- scores_v22$scores[,"TK"]
data_emne2_v22$PCK <- scores_v22$scores[,"PCK"]
data_emne2_v22$TPACK <- scores_v22$scores[,"TPACK"]

#describe(data_emne2_v22[data_emne2_v22$datasett == "pre",])
scores_v22_pre <- scoreItems(TPACK_PST_emne2_v22,
                             data_emne2_v22[data_emne2_v22$datasett == "pre" &
                                            data_emne2_v22$Subject %in%
                                              stud_alle_v22,])
describe(scores_v22_pre$scores)

#describe(data_emne2_v22[data_emne2_v22$datasett == "post",])
scores_v22_post <- scoreItems(TPACK_PST_emne2_v22,
                              data_emne2_v22[data_emne2_v22$datasett == "post" &
                                             data_emne2_v22$Subject %in%
                                               stud_alle_v22,])
describe(scores_v22_post$scores)
```

```{r create_table_v22}
#!NB Sjekk om rekkefølgen her er riktig? Eller alternativt bruk data_emne2_v22-dataframen

# gjør om til loop

# pre
TK_pre <- scores_v22_pre$scores[,"TK"]
TK_alpha_pre <- scores_v22_pre$corrected["TK","TK"]

PCK_pre <- scores_v22_pre$scores[,"PCK"]
PCK_alpha_pre <- scores_v22_pre$corrected["PCK","PCK"]

TPACK_pre <- scores_v22_pre$scores[,"TPACK"]
TPACK_alpha_pre <- scores_v22_pre$corrected["TPACK","TPACK"]

#post
TK_post <- scores_v22_post$scores[,"TK"]
TK_alpha_post <- scores_v22_post$corrected["TK","TK"]

PCK_post <- scores_v22_post$scores[,"PCK"]
PCK_alpha_post <- scores_v22_post$corrected["PCK","PCK"]

TPACK_post <- scores_v22_post$scores[,"TPACK"]
TPACK_alpha_post <- scores_v22_post$corrected["TPACK","TPACK"]

# cohens d
TK_cd <- effsize::cohen.d(TK_post, TK_pre, 
                 hedges.correction=TRUE, within=T, paired=T, noncentral=T)#, conf.level=0.80)
PCK_cd <- effsize::cohen.d(PCK_post, PCK_pre, 
                  hedges.correction=TRUE, within=T, paired=T, noncentral=T)#, conf.level=0.80)
TPACK_cd <- effsize::cohen.d(TPACK_post, TPACK_pre, 
                    hedges.correction=TRUE, within=T, paired=T, noncentral=T)#, conf.level=0.80)

# tabell
TK_rad <- c(mean(TK_pre), mean(TK_post), TK_cd$estimate, TK_cd$conf.int, TK_alpha_pre, TK_alpha_post)
PCK_rad <- c(mean(PCK_pre), mean(PCK_post), PCK_cd$estimate, PCK_cd$conf.int, PCK_alpha_pre, PCK_alpha_post)
TPACK_rad <- c(mean(TPACK_pre), mean(TPACK_post), TPACK_cd$estimate, TPACK_cd$conf.int, TPACK_alpha_pre, TPACK_alpha_post)
tabell_v22 <- t(data.frame(TK_rad, #TCK_rad, TPK_rad, 
                           PCK_rad, TPACK_rad))
rownames(tabell_v22) <- c("TK", "PCK", "TPACK")
colnames(tabell_v22) <- c("pre","post", "cohen's d", "ci lower", "ci upper", "alpha_pre", "alpha_post")

pander(tabell_v22)
```

Er det noen signifikant forskjell mellom de ulike periodene? 

sjekk forutsetninger for t-test (normalitet og lik varians)

```{r}
# sjekk av antagelser
# ref. Pearce and Derrick (2019)

TK22_pre <- data_emne2_v22[alle_studenter_v22 & data_emne2_v22$datasett == "pre", "TK"]
TK22_post <- data_emne2_v22[alle_studenter_v22 & data_emne2_v22$datasett == "post", "TK"]
PCK22_pre <- data_emne2_v22[alle_studenter_v22 & data_emne2_v22$datasett == "pre", "PCK"]
PCK22_post <- data_emne2_v22[alle_studenter_v22 & data_emne2_v22$datasett == "post", "PCK"]
TPACK22_pre <- data_emne2_v22[alle_studenter_v22 & data_emne2_v22$datasett == "pre", "TPACK"]
TPACK22_post <- data_emne2_v22[alle_studenter_v22 & data_emne2_v22$datasett == "post", "TPACK"]

df22_prepost <- data.frame(TK = c(TK22_pre, TK22_post),
                    PCK = c(PCK22_pre, PCK22_post),
                    TPACK = c(TPACK22_pre, TPACK22_post),
                    prototype = c(rep("pre", length(TK22_pre)), 
                                  rep("post", length(TK22_post))))

```

```{r}
# Levene's (Mean) test for equal variances
# ref https://www.statology.org/levenes-test-r/
#library(car)
leveneTest(TK ~ prototype, center=mean, data = df22_prepost)
leveneTest(PCK ~ prototype, center=mean, data = df22_prepost)
leveneTest(TPACK ~ prototype, center=mean, data = df22_prepost)
```

Ulik varians.

```{r}
ks.test(TK22_pre, "pnorm")
ks.test(TK22_post, "pnorm")
ks.test(PCK22_pre, "pnorm")
ks.test(PCK22_post, "pnorm")
ks.test(TPACK22_pre, "pnorm")
ks.test(TPACK22_post, "pnorm")
```

Ikke normalfordelt.

```{r}
# Yuen-Welch test
# ref.: Mair & Wilcox, 2020
#library(WRS2)
yuend(TK22_pre, TK22_post)
yuend(PCK22_pre, PCK22_post)
yuend(TPACK22_pre, TPACK22_post)
```

```{r}
t.test(data_emne2_v22[alle_studenter_v22 & data_emne2_v22$datasett=="pre", "TK"],
       data_emne2_v22[alle_studenter_v22 & data_emne2_v22$datasett=="post", "TK"], 
       paired=T)

t.test(data_emne2_v22[alle_studenter_v22 & data_emne2_v22$datasett=="pre", "PCK"],
       data_emne2_v22[alle_studenter_v22 & data_emne2_v22$datasett=="post", "PCK"], 
       paired=T)

t.test(data_emne2_v22[alle_studenter_v22 & data_emne2_v22$datasett=="pre", "TPACK"],
       data_emne2_v22[alle_studenter_v22 & data_emne2_v22$datasett=="post", "TPACK"], 
       paired=T)
```

Se på forandring i TPACK

```{r calc_diff_v22}
calc_diff_TPACK_v22 <- function (TPACK_element) {

df <- data_emne2_v22[data_emne2_v22$datasett == "pre" | 
                     data_emne2_v22$datasett == "post",
                     c("Subject","datasett",TPACK_element)] # "group",

# sorter på datasett
df <-  df[
                          with(df, 
                          order(datasett,Subject, # rekkefølge
                          method="radix"#, # for faktorer, brukes med c nedenfor
                          #decreasing = c(-1,0))
                          ))
                          ,] # bare datasett (1) 

# regn ut TPACK pre/post for hver person
ifelse (length(TPACK_element) > 1, {
TPACK_pre <- rowMeans(df[df$datasett=="pre",3:ncol(df)],na.rm=TRUE)
TPACK_post <- rowMeans(df[df$datasett=="post",3:ncol(df)],na.rm=TRUE)
},
{
  TPACK_pre <- df[df$datasett=="pre",TPACK_element]
  TPACK_post <- df[df$datasett=="post",TPACK_element]
})
TPACK_merge <- c(TPACK_pre,TPACK_post)

# legg til resultat
df$TPACK_pre_post <- TPACK_merge

# sorter på person; 
df <-  df[
                          with(df, 
                          order(Subject,datasett, # rekkefølge
                          method="radix"))#, # for faktorer, brukes med c nedenfor
                          #decreasing = c(0,1)))
                          ,] # bare datasett (1) 

# behold bare personer som er med i  både pre/post
df_filtered <- df[!df$Subject %in% c("Benjamin","Gabriel","Olava"),]

# beregn differanse; pre inneholder riktig diff-verdi
df_filtered$diff <-
  c(diff(df_filtered[,ncol(df_filtered)]),0)

diff_TPACK <- df_filtered[df_filtered$datasett=="pre",c(1,ncol(df_filtered))]

return (diff_TPACK)
}

TK_diff_v22 <- calc_diff_TPACK_v22(TK_PST22_pos)
PCK_diff_v22 <- calc_diff_TPACK_v22(PCK_PST22_pos)
TPACK_diff_v22 <- calc_diff_TPACK_v22(TPACK_PST22_pos)

# lag dataframe
diff_full_v22 <- data.frame("Subject"=TK_diff_v22[,1], #"group" = TK_diff[,"group"],
  "TK"=TK_diff_v22[,2],
  "PCK"=PCK_diff_v22[,2],
  "TPACK"=TPACK_diff_v22[,2])
```

##### reliable change in individual scores

beregn SID (Estrada et al., 2019).

```{r}
SID <- TK_diff_v22[,2] / SD(TK_diff_v22[,2])

# percentage of reliable improvements
pri <- sum(SID > 1.645, na.rm=TRUE)
pri
pri / length(SID)
```

```{r}
SID <- PCK_diff_v22[,2] / SD(PCK_diff_v22[,2])

# percentage of reliable improvements
pri <- sum(SID > 1.645, na.rm=TRUE)
pri
pri / length(SID)
```

```{r}
SID <- TPACK_diff_v22[,2] / SD(TPACK_diff_v22[,2])

# percentage of reliable improvements
pri <- sum(SID > 1.645, na.rm=TRUE)
pri
pri / length(SID)
```

##### Sjekk for RTM

RTM = regression to the mean

Ref.: Marsden og Torgerson (2012)

```{r}
#plot diff vs pre
ggplot(data.frame("postpre" = diff_full_v22[,"TK"], 
                  "TK" = data_emne2_v22[data_emne2_v22$datasett == "pre" & 
                                        data_emne2_v22$Subject %in% stud_alle_v22,"TK"]),
       aes(x=TK,y=postpre)) +
  geom_jitter()  +
  stat_smooth(method = "lm", formula = 'y ~ x', se=F,fullrange = F)
```

```{r}
#plot diff vs pre
ggplot(data.frame("postpre" = diff_full_v22[,"PCK"], 
                  "PCK" = data_emne2_v22[data_emne2_v22$datasett == "pre" & 
                                        data_emne2_v22$Subject %in% stud_alle_v22,"PCK"]),
       aes(x=PCK,y=postpre)) +
  geom_jitter()  +
  stat_smooth(method = "lm", formula = 'y ~ x', se=F,fullrange = F)
```

```{r}
#plot diff vs pre
ggplot(data.frame("postpre" = diff_full_v22[,"TPACK"], 
                  "TPACK" = data_emne2_v22[data_emne2_v22$datasett == "pre" & 
                                        data_emne2_v22$Subject %in% stud_alle_v22,"TPACK"]),
       aes(x=TPACK,y=postpre)) +
  geom_jitter()  +
  stat_smooth(method = "lm", formula = 'y ~ x', se=F,fullrange = F)
```

#### Prototype 2

```{r descriptive_stat_v23}
scores_v23 <- scoreItems(TPACK_PST_emne2_v23, data_emne2_v23)
data_emne2_v23$TK <- scores_v23$scores[,"TK"]
data_emne2_v23$PCK <- scores_v23$scores[,"PCK"]
data_emne2_v23$TPACK <- scores_v23$scores[,"TPACK"]

#describe(data_emne2_v23[data_emne2_v23$datasett == "pre",])
scores_v23_pre <- scoreItems(TPACK_PST_emne2_v23, data_emne2_v23[data_emne2_v23$datasett == "pre1" & data_emne2_v23$Subject %in% stud_alle_v23,])
describe(scores_v23_pre$scores)

#describe(data_emne2_v23[data_emne2_v23$datasett == "post",])
scores_v23_post <- scoreItems(TPACK_PST_emne2_v23,
                              data_emne2_v23[data_emne2_v23$datasett == "post" &
                                             data_emne2_v23$Subject %in%
                                               stud_alle_v23,])
describe(scores_v23_post$scores)
```

```{r create_table_v23}
#!NB Sjekk om rekkefølgen her er riktig? Eller alternativt bruk data_emne2_v23-dataframen

# gjør om til loop

# pre
TK_pre <- scores_v23_pre$scores[,"TK"]
TK_alpha_pre <- scores_v23_pre$corrected["TK","TK"]

PCK_pre <- scores_v23_pre$scores[,"PCK"]
PCK_alpha_pre <- scores_v23_pre$corrected["PCK","PCK"]

TPACK_pre <- scores_v23_pre$scores[,"TPACK"]
TPACK_alpha_pre <- scores_v23_pre$corrected["TPACK","TPACK"]

#post
TK_post <- scores_v23_post$scores[,"TK"]
TK_alpha_post <- scores_v23_post$corrected["TK","TK"]

PCK_post <- scores_v23_post$scores[,"PCK"]
PCK_alpha_post <- scores_v23_post$corrected["PCK","PCK"]

TPACK_post <- scores_v23_post$scores[,"TPACK"]
TPACK_alpha_post <- scores_v23_post$corrected["TPACK","TPACK"]

# cohens d
TK_cd <- effsize::cohen.d(TK_post, TK_pre, 
                 hedges.correction=TRUE, within=T, paired=T, noncentral=T)#, conf.level=0.80)
PCK_cd <- effsize::cohen.d(PCK_post, PCK_pre, 
                  hedges.correction=TRUE, within=T, paired=T, noncentral=T)#, conf.level=0.80)
TPACK_cd <- effsize::cohen.d(TPACK_post, TPACK_pre, 
                    hedges.correction=TRUE, within=T, paired=T, noncentral=T)#, conf.level=0.80)

# tabell
TK_rad <- c(mean(TK_pre), mean(TK_post), TK_cd$estimate, TK_cd$conf.int, TK_alpha_pre, TK_alpha_post)
PCK_rad <- c(mean(PCK_pre), mean(PCK_post), PCK_cd$estimate, PCK_cd$conf.int, PCK_alpha_pre, PCK_alpha_post)
TPACK_rad <- c(mean(TPACK_pre), mean(TPACK_post), TPACK_cd$estimate, TPACK_cd$conf.int, TPACK_alpha_pre, TPACK_alpha_post)
tabell_v23 <- t(data.frame(TK_rad, #TCK_rad, TPK_rad, 
                           PCK_rad, TPACK_rad))
rownames(tabell_v23) <- c("TK", "PCK", "TPACK")
colnames(tabell_v23) <- c("pre","post", "cohen's d", "ci lower", "ci upper", "alpha_pre", "alpha_post")

pander(tabell_v23)
```

Er det noen signifikant forskjell mellom de ulike periodene? 

sjekk forutsetninger for t-test (normalitet og lik varians)

```{r}
# sjekk av antagelser
# ref. Pearce and Derrick (2019)

TK23_pre <- data_emne2_v23[alle_studenter_v23 & data_emne2_v23$datasett == "pre1", "TK"]
TK23_post <- data_emne2_v23[alle_studenter_v23 & data_emne2_v23$datasett == "post", "TK"]
PCK23_pre <- data_emne2_v23[alle_studenter_v23 & data_emne2_v23$datasett == "pre1", "PCK"]
PCK23_post <- data_emne2_v23[alle_studenter_v23 & data_emne2_v23$datasett == "post", "PCK"]
TPACK23_pre <- data_emne2_v23[alle_studenter_v23 & data_emne2_v23$datasett == "pre1", "TPACK"]
TPACK23_post <- data_emne2_v23[alle_studenter_v23 & data_emne2_v23$datasett == "post", "TPACK"]

df23_prepost <- data.frame(TK = c(TK23_pre, TK23_post),
                    PCK = c(PCK23_pre, PCK23_post),
                    TPACK = c(TPACK23_pre, TPACK23_post),
                    prototype = c(rep("pre1", length(TK23_pre)), 
                                  rep("post", length(TK23_post))))

```

```{r}
# Levene's (Mean) test for equal variances
# ref https://www.statology.org/levenes-test-r/
#library(car)
leveneTest(TK ~ prototype, center=mean, data = df23_prepost)
leveneTest(PCK ~ prototype, center=mean, data = df23_prepost)
leveneTest(TPACK ~ prototype, center=mean, data = df23_prepost)
```

Ulik varians.

```{r}
ks.test(TK23_pre, "pnorm")
ks.test(TK23_post, "pnorm")
ks.test(PCK23_pre, "pnorm")
ks.test(PCK23_post, "pnorm")
ks.test(TPACK23_pre, "pnorm")
ks.test(TPACK23_post, "pnorm")
```

Ikke normalfordelt.

```{r}
# Wilcoxon signed-rank test (Mann-Whitney)
wilcox.test(TK23_pre, TK23_post, paired=T)
wilcox.test(PCK23_pre, PCK23_post, paired=T)
wilcox.test(TPACK23_pre, TPACK23_post, paired=T)
```

```{r}
# Yuen-Welch test
# ref.: Mair & Wilcox, 2020
#library(WRS2)
yuend(TK23_pre, TK23_post)
yuend(PCK23_pre, PCK23_post)
yuend(TPACK23_pre, TPACK23_post)
```

```{r}
t.test(data_emne2_v23[alle_studenter_v23 & data_emne2_v23$datasett=="pre1", "TK"],
       data_emne2_v23[alle_studenter_v23 & data_emne2_v23$datasett=="post", "TK"], 
       paired=T)

t.test(data_emne2_v23[alle_studenter_v23 & data_emne2_v23$datasett=="pre1", "PCK"],
       data_emne2_v23[alle_studenter_v23 & data_emne2_v23$datasett=="post", "PCK"], 
       paired=T)

t.test(data_emne2_v23[alle_studenter_v23 & data_emne2_v23$datasett=="pre1", "TPACK"],
       data_emne2_v23[alle_studenter_v23 & data_emne2_v23$datasett=="post", "TPACK"], 
       paired=T)
```

Se på forandring i TPACK

```{r calc_diff_v23}
calc_diff_TPACK_v23 <- function (TPACK_element) {

df <- data_emne2_v23[data_emne2_v23$datasett == "pre1" | 
                     data_emne2_v23$datasett == "post",
                     c("Subject","datasett",TPACK_element)] # "group",

# sorter på datasett
df <-  df[
                          with(df, 
                          order(datasett,Subject, # rekkefølge
                          method="radix"#, # for faktorer, brukes med c nedenfor
                          #decreasing = c(-1,0))
                          ))
                          ,] # bare datasett (1) 

# regn ut TPACK pre/post for hver person
ifelse (length(TPACK_element) > 1, {
TPACK_pre <- rowMeans(df[df$datasett=="pre1",3:ncol(df)],na.rm=TRUE)
TPACK_post <- rowMeans(df[df$datasett=="post",3:ncol(df)],na.rm=TRUE)
},
{
  TPACK_pre <- df[df$datasett=="pre1",TPACK_element]
  TPACK_post <- df[df$datasett=="post",TPACK_element]
})
TPACK_merge <- c(TPACK_pre,TPACK_post)

# legg til resultat
df$TPACK_pre_post <- TPACK_merge

# sorter på person; 
df <-  df[
                          with(df, 
                          order(Subject,datasett, # rekkefølge
                          method="radix"))#, # for faktorer, brukes med c nedenfor
                          #decreasing = c(0,1)))
                          ,] # bare datasett (1) 

# behold bare personer som er med i  både pre/post
df_filtered <- df[df$Subject %in% stud_alle_v23,]

# beregn differanse; pre inneholder riktig diff-verdi
df_filtered$diff <-
  c(diff(df_filtered[,ncol(df_filtered)]),0)

diff_TPACK <- df_filtered[df_filtered$datasett=="pre1",c(1,ncol(df_filtered))]

return (diff_TPACK)
}

TK_diff_v23 <- calc_diff_TPACK_v23(TK_PST23_pos)
PCK_diff_v23 <- calc_diff_TPACK_v23(PCK_PST23_pos)
TPACK_diff_v23 <- calc_diff_TPACK_v23(TPACK_PST23_pos)

# lag dataframe
diff_full_v23 <- data.frame("Subject"=TK_diff_v23[,1], #"group" = TK_diff[,"group"],
  "TK"=TK_diff_v23[,2],
  "PCK"=PCK_diff_v23[,2],
  "TPACK"=TPACK_diff_v23[,2])
```

##### reliable change in individual scores

beregn SID (Estrada et al., 2019).

```{r}
SID <- TK_diff_v23[,2] / SD(TK_diff_v23[,2])

# percentage of reliable improvements
pri <- sum(SID > 1.645, na.rm=TRUE)
pri
pri / length(SID)
```

```{r}
SID <- PCK_diff_v23[,2] / SD(PCK_diff_v23[,2])

# percentage of reliable improvements
pri <- sum(SID > 1.645, na.rm=TRUE)
pri
pri / length(SID)
```

```{r}
SID <- TPACK_diff_v23[,2] / SD(TPACK_diff_v23[,2])

# percentage of reliable improvements
pri <- sum(SID > 1.645, na.rm=TRUE)
pri
pri / length(SID)
```

##### Sjekk for RTM

RTM = regression to the mean

Ref.: Marsden og Torgerson (2012)

```{r}
#plot diff vs pre
ggplot(data.frame("postpre" = diff_full_v23[,"TK"], 
                  "TK" = data_emne2_v23[data_emne2_v23$datasett == "pre1" & 
                                        data_emne2_v23$Subject %in% stud_alle_v23,"TK"]),
       aes(x=TK,y=postpre)) +
  geom_jitter()  +
  stat_smooth(method = "lm", formula = 'y ~ x', se=F,fullrange = F)
```

```{r}
#plot diff vs pre
ggplot(data.frame("postpre" = diff_full_v23[,"PCK"], 
                  "PCK" = data_emne2_v23[data_emne2_v23$datasett == "pre1" & 
                                        data_emne2_v23$Subject %in% stud_alle_v23,"PCK"]),
       aes(x=PCK,y=postpre)) +
  geom_jitter()  +
  stat_smooth(method = "lm", formula = 'y ~ x', se=F,fullrange = F)
```

```{r}
#plot diff vs pre
ggplot(data.frame("postpre" = diff_full_v23[,"TPACK"], 
                  "TPACK" = data_emne2_v23[data_emne2_v23$datasett == "pre1" & 
                                        data_emne2_v23$Subject %in% stud_alle_v23,"TPACK"]),
       aes(x=TPACK,y=postpre)) +
  geom_jitter()  +
  stat_smooth(method = "lm", formula = 'y ~ x', se=F,fullrange = F)
```


### Fixed-effects

Before doing the regression analysis, we need to decide how to model the data. Should we model it through pooled OLS, fixed effects or random effects? 

... Legg til sjekker av hvilken modell som bør brukes

```{r}
within22 <- plm(TPACK ~ TK + PCK + factor(Subject) + factor(datasett), 
              data=data_emne2_v22,
              index=c("Subject","datasett"), model="within")
firstdifference22 <- plm(TPACK ~ TK + PCK + factor(Subject),# + factor(datasett),
              data=data_emne2_v22,
              index=c("Subject","datasett"), model="fd")
```

```{r within_3pt}
within23 <- plm(TPACK ~ TK + PCK + factor(Subject) + factor(datasett), 
              data=data_emne2_v23,
              index=c("Subject","datasett"), model="within")
firstdifference23 <- plm(TPACK ~ TK + PCK + factor(Subject),# + factor(datasett),
              data=data_emne2_v23,
              index=c("Subject","datasett"), model="fd")
```

```{r within_2pt}
within23_prepost <- plm(TPACK ~ TK + PCK + factor(Subject) + factor(datasett), 
              data=data_emne2_v23[data_emne2_v23$datasett == "pre1" | data_emne2_v23$datasett == "post",],
              index=c("Subject","datasett"), model="within")
firstdifference23_prepost <- plm(TPACK ~ TK + PCK + factor(Subject),# + factor(datasett),
              data=data_emne2_v23[data_emne2_v23$datasett == "pre1" | data_emne2_v23$datasett == "post",],
              index=c("Subject","datasett"), model="fd")
```

```{r}
screenreg(list("within22" = within22,
               "fd22" = firstdifference22,
               "within23" = within23,
               "fd23" = firstdifference23,
               "within23pp" = within23_prepost,
               "fd23pp" = firstdifference23_prepost))
```

### Within-between models

To calculate the within-between model for prototype 2, all three datapoints were needed. This was not done in paper 4. 

Prototype 1

prepare centred and demeaned variabels

create a dummy variable for each individual
or
gruppesentrering, _S

```{r prot1_centering}
# TK
# calculate mean for each subject / id; ref.: https://stackoverflow.com/a/34502529
data_emne2_v22$TK_mean <- with(data_emne2_v22, ave(x = TK, Subject, FUN = mean))
data_emne2_v22$TK_S <- data_emne2_v22$TK - data_emne2_v22$TK_mean

# PCK interval
data_emne2_v22$PCK_mean <- with(data_emne2_v22, ave(x = PCK, Subject, FUN = mean))
data_emne2_v22$PCK_S <- data_emne2_v22$PCK - data_emne2_v22$PCK_mean

# TPACK
data_emne2_v22$TPACK_mean <- with(data_emne2_v22, ave(x = TPACK, Subject, FUN = mean))
data_emne2_v22$TPACK_S <- data_emne2_v22$TPACK - data_emne2_v22$TPACK_mean
```

```{r prot2_centering}
# TK
# calculate mean for each subject / id; ref.: https://stackoverflow.com/a/34502529
data_emne2_v23$TK_mean <- with(data_emne2_v23, ave(x = TK, Subject, FUN = mean))
data_emne2_v23$TK_S <- data_emne2_v23$TK - data_emne2_v23$TK_mean

# PCK interval
data_emne2_v23$PCK_mean <- with(data_emne2_v23, ave(x = PCK, Subject, FUN = mean))
data_emne2_v23$PCK_S <- data_emne2_v23$PCK - data_emne2_v23$PCK_mean

# TPACK
data_emne2_v23$TPACK_mean <- with(data_emne2_v23, ave(x = TPACK, Subject, FUN = mean))
data_emne2_v23$TPACK_S <- data_emne2_v23$TPACK - data_emne2_v23$TPACK_mean
```

Hybrid with variable coefficients was chosen since there was only one significant dependent variable in the fixed-effects model (Bell et al., 2019).

Ville det vært bedre å sammenligne med en enkel hybrid model uten variable koeffisienter først, for mer direkte sammenligning med FE-modellen?

```{r eval=T}
# TK_S within effects
# TK_mean between effects
# factor(datasett) fixed time effects
# TK_S|id variable coefficient TK_S per individual

hybrid_TK <- lme(TPACK ~ TK_S + TK_mean + PCK_S + PCK_mean + factor(datasett), 
                   random=~1|Subject, method="REML", data=data_emne2_v22[alle_studenter_v22,])

summary(hybrid_TK)
```

bare pre-post

```{r eval=T}
# TK_S within effects
# TK_mean between effects
# factor(datasett) fixed time effects
# TK_S|id variable coefficient TK_S per individual

hybrid_TK <- lme(TPACK ~ TK_S + TK_mean + PCK_S + PCK_mean + factor(datasett), 
                   random=~TK_S|Subject, method="REML", data=data_emne2_v22[alle_studenter_v22,])

summary(hybrid_TK)
```

alle studentene

```{r eval=T}
# TK_S within effects
# TK_mean between effects
# factor(datasett) fixed time effects
# TK_S|id variable coefficient TK_S per individual

hybrid_TK <- lme(TPACK ~ TK_S + TK_mean + PCK_S + PCK_mean + factor(datasett), 
                   random=~TK_S|Subject, method="REML", data=data_emne2_v22)

summary(hybrid_TK)
```

#### Comparison hybrid models

Use FIML (ML) for model comparison, REML for parameter estimates (Christophersen, 2018).

Prototype 1

```{r eval=T}
# TK_S within effects
# TK_mean between effects
# factor(datasett) fixed time effects
# TK_S|id variable coefficient TK_S per individual

prot1_within_between <- lme(TPACK ~ TK_S + TK_mean + PCK_S + PCK_mean + factor(datasett), 
                   random=~1|Subject, method="ML", data=data_emne2_v22[alle_studenter_v22,])

prot1_var_coeff <- lme(TPACK ~ TK_S + TK_mean + PCK_S + PCK_mean + factor(datasett), 
                   random=~TK_S|Subject, method="ML", data=data_emne2_v22[alle_studenter_v22,])
```

Prototype 2

```{r prot2_wb, eval=T}
# TK_S within effects
# TK_mean between effects
# factor(datasett) fixed time effects
# TK_S|id variable coefficient TK_S per individual

# only 13 from both pre-post
prot2_within_between_13 <- lme(TPACK ~ TK_S + TK_mean + PCK_S + PCK_mean + factor(datasett), 
                   random=~1|Subject, method="REML", 
                   data=data_emne2_v23[alle_studenter_v23 & !data_emne2_v23$datasett == "pre2",])

# all 30 for comparison with var coeff below
prot2_within_between <- lme(TPACK ~ TK_S + TK_mean + PCK_S + PCK_mean + factor(datasett), 
                   random=~1|Subject, method="ML", data=data_emne2_v23[ !data_emne2_v23$datasett == "pre2",]) # [alle_studenter_v23,]
```

```{r prot2_vc}
# does not converge with only 13 respondents
prot2_var_coeff <- lme(TPACK ~ TK_S + TK_mean + PCK_S + PCK_mean + factor(datasett), 
                   random=~TK_S|Subject, method="ML", data=data_emne2_v23[ !data_emne2_v23$datasett == "pre2",]) # alle_studenter_v23 &
```

```{r}
# print summary of models
screenreg(list("P1_wb" = prot1_within_between, 
               "P1_vc" = prot1_var_coeff,
               "P2_wb" = prot2_within_between, 
               "P2_vc" = prot2_var_coeff,
               "P2_13" = prot2_within_between_13))
```

chi-square comparison, significant difference between models?

```{r}
anova(prot1_within_between, prot1_var_coeff)

anova(prot2_within_between, prot2_var_coeff)
```

AICc for small samples

```{r}
library(MuMIn) # AICc
AICc(prot1_within_between, prot1_var_coeff, REML=F)
AICc(prot2_within_between, prot2_var_coeff, REML=F)
```

summary of details for best fit models

```{r}
prot1_var_coeff_REML <- lme(TPACK ~ TK_S + TK_mean + PCK_S + PCK_mean + factor(datasett), 
                   random=~TK_S|Subject, method="REML", data=data_emne2_v22[alle_studenter_v22,])

prot2_within_between_13_REML <- lme(TPACK ~ TK_S + TK_mean + PCK_S + PCK_mean + factor(datasett), 
                   random=~1|Subject, method="REML", 
                   data=data_emne2_v23[alle_studenter_v23 & !data_emne2_v23$datasett == "pre2",])

summary(prot1_var_coeff_REML)
summary(prot2_within_between_13_REML)
```


# Results

